\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-08 jasa/paper-jasa.tex
Improve first example

As a simple and illustrative example, let us first consider the approximation of
the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the 1D mixtures
$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
Figure~\ref{fig:1a} and defined as
\begin{align}
p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - \gamma) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \gamma) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}), \\
p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
\end{align}
where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $\gamma$ is a parameter of $p(\mathbf{x}|\theta_0)$
set to $0.05$. Samples drawn from $p(\mathbf{x}|\theta_0)$ are shown in the
figure and will be used later for inference.

Using the method outlined in Section~\ref{sec:approx},
figures~\ref{fig:1b}, \ref{fig:1c} and \ref{fig:1d} show the approximated
ratio when using respectively a linear model, a 2-layer neural network and a random
forest. The blue curve corresponds to the true target log-ratio, the green curve is
the direct approximation $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$
without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
s(\mathbf{x}) | \theta_0)}{\hat p(\hat s(\mathbf{x}) | \theta_1)}$ as calibrated
using histograms. Finally, the
cyan curve shows the approximated ratio when both decomposing the mixtures and calibrating the underlying classifiers, as
described in Section~\ref{sec:mixture}.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1a.pdf}
        \caption{$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$}
        \label{fig:1a}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
        \caption{Linear model}
        \label{fig:1b}
    \end{subfigure}

@@

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1c.pdf}
        \caption{Neural network}
        \label{fig:1c}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1d.pdf}
        \caption{Random forest}
        \label{fig:1d}
    \end{subfigure}
    \caption{Approximation of the log-likelihood ratio
             $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$
             with calibrated classifiers.}
\end{figure}

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat
s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate
estimates of the ratio, even when using universal approximators such as a neural
network or a random forest. This observation is at the origin of this work, and
led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
forming the equivalent statistic
$\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\theta_0)}{\hat p(\hat
s(\mathbf{x})|\theta_1)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be more accurate for all three classifiers, even
when the direct approximation was not very good in the first place, e.g. as for the
linear model. In regions where the densities are low (i.e., at the
boundaries) the approximated ratio show to be less acurate since the underlying
classifier cannot make precise predictions, because of the few training samples
it is trained on in these parts of the input space. When leveraging the fact
that densities are mixtures and decomposing their ratio, the capacity of the
underlying classifiers can be focused on easier supervised learning tasks,
resulting as expected in even more accurate approximations, as shown in cyan.

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work best, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$ which is
very close to the optimal decision function. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis.

@@

    \begin{subfigure}[t]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2a.pdf}
        \caption{True MLEs vs. approximated MLEs.}
        \label{fig:2a}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2b.pdf}
        \caption{$-2 \log \frac{p({\cal D}|\gamma=0.05)}{\sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)}$}
        \label{fig:2b}
    \end{subfigure}

@@

Let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0$ versus the alternative $H_1: \gamma > 0$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | \gamma = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)} \; .
\end{equation}
Following the approach outlined in
Section~\ref{sec:generalized-likelihood-ratio}, computing the generalized
likelihood ratio statistic requires finding the maximum likelihood estimator
(MLE) for the denominator, which can be carried out using approximated
likelihood ratios. Reusing the decomposed neural network-based approximation and
minimizing the mean negative log-ratio of some observed data generated for
$\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. The figure clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating estimator bias. As a second check, Figure \ref{fig:2b} also confirms
that the generalized likelihood ratio statistic obtained with our method is $\chi^2$ distributed,
as should be the true statistic.

>>>>

As a simple and illustrative example, let us first consider the approximation of
the likelihood log-ratio $\log \left( r(\mathbf{x};\gamma=0.05,\gamma=0) \right)$ between the 1D mixtures
$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$ illustrated in
Figure~\ref{fig:1a} and defined as
\begin{align}
p(\mathbf{x}|\gamma) &= (\frac{1}{2} - \frac{\gamma}{2}) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \frac{\gamma}{2}) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}),
\end{align}
where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$. Samples drawn for the nominal value $\gamma=0.05$ are shown in the
figure and will be used later for inference.

Using the method outlined in Section~\ref{sec:approx},
figures~\ref{fig:1b}, \ref{fig:1c} and \ref{fig:1d} show the approximated
ratio when using respectively a linear model, a 2-layer neural network and a random
forest. The blue curve corresponds to the true target log-ratio, the green curve is
the direct approximation $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$
without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
s(\mathbf{x}) | \gamma=0.05)}{\hat p(\hat s(\mathbf{x}) | \gamma=0)}$ as calibrated
using histograms. Finally, the
cyan curve shows the approximated ratio when both decomposing the mixtures and calibrating the underlying classifiers, as
described in Section~\ref{sec:mixture}.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1a.pdf}
        \caption{$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$}
        \label{fig:1a}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
        \caption{Linear model}
        \label{fig:1b}
    \end{subfigure}

@@

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1c.pdf}
        \caption{Neural network}
        \label{fig:1c}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1d.pdf}
        \caption{Random forest}
        \label{fig:1d}
    \end{subfigure}
    \caption{Approximation of the log-likelihood ratio
             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
             with calibrated classifiers.}
\end{figure}

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat
s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate
estimates of the ratio, even when using universal approximators such as a neural
network or a random forest. This observation is at the origin of this work, and
led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
forming the equivalent statistic
$\frac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat
s(\mathbf{x})|\gamma=0)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be more accurate for all three classifiers, even
when the direct approximation was not very good in the first place, e.g. as for the
linear model. In regions where the densities are low (i.e., at the
boundaries) the approximated ratio show to be less acurate since the underlying
classifier cannot make precise predictions, because of the few training samples
it is trained on in these parts of the input space. When leveraging the fact
that densities are mixtures and decomposing their ratio, the capacity of the
underlying classifiers can be focused on easier supervised learning tasks,
resulting as expected in even more accurate approximations, as shown in cyan.

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work best, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$ which is
very close to the optimal decision function. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis. Overall, this first example confirms
that even if the classifier does a poor joob at modeling the optimal decision
function, good approximations of the likelihood ratio can still be obtained
with the proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
the mixture when possible.

@@

    \begin{subfigure}[t]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2a.pdf}
        \caption{True MLEs vs. approximated MLEs.}
        \label{fig:2a}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2b.pdf}
        \caption{$-2 \log \frac{p({\cal D}|\gamma=0.05)}{\sup_{\gamma \in \Theta} p({\cal D} | \gamma)}$}
        \label{fig:2b}
    \end{subfigure}

@@

Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
\end{equation}
By exploiting the fact that
\begin{equation}
\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
\end{equation}
the generalized likelihood ratio statistic can be computed by evaluating
both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
In particular, the maximum likelihood estimator (MLE) in the second term
can be found as outlined in
Section~\ref{sec:generalized-likelihood-ratio}.
Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. It clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating bias. Figure \ref{fig:2b} also confirms
that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
as should be the true statistic when the null is true.
